---
title: "Marketing Spend and Purchase Amounts: A Campaign Analysis"
author: "Prashorst Meas"
format:
  html: default
  pdf:     
    fig-pos: "H"
    include-in-header:
      text: |
        \usepackage{float}
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
#| echo: false
#| message: false
#| warning: false

library(readr)
library(dplyr)
library(tidyr)
library(lubridate)
library(ggplot2)
library(knitr)
```

```{r}
#| label: clean-data
#| echo: false


marketing_raw <- read_csv(
  "data/marketing_data.csv",
  na = c("", "N/A"),
  show_col_types = FALSE
)

marketing <- marketing_raw |>
  mutate(
    signup_date = as_date(parse_date_time(signup_date,
                                          orders = c("dmy", "mdy"),
                                          quiet = TRUE)),
    segment = factor(segment,
                     levels = c("Consumer", "SMB", "Enterprise")),
    notes = factor(replace_na(as.character(notes), "none"),
                   levels = c("none", "contacted",
                              "no response", "returned"))
  ) |>
  filter(!is.na(purchase_amount))
```

# Introduction

Online retailers often use personalised marketing strategies to encourage customers to buy more. However, it is important for businesses to know whether spending more money on advertising per customer actually leads to higher sales. This report looks at data from a recent personalised marketing campaign to understand how customer behaviour changed based on the amount spent on them. Understanding this relationship helps businesses decide how to best allocate their marketing budgets in the future.

# Research Question

This report aims to answer whether a higher marketing spend per customer results in a higher purchase amount. This is an important question because it helps determine whether the personalised marketing strategy is profitable and should be continued. If there is a strong positive link, the business might increase marketing spend; if not, it might need to try cheaper advertising methods.

# Dataset

The dataset records information from a recent marketing campaign, with one row for every customer contacted. It includes details such as marketing spend, site visits, purchase amounts, customer segment, and outreach notes. To clean the data, I first standardised the signup dates that were in different formats and treated empty strings and "N/A" text as missing values. I then replaced missing values in the notes column with "none" so it could be used as a proper category. Finally, I removed any rows that were missing the purchase amount, as this is the main outcome variable needed to answer my research question.

## Dataset Description

The cleaned marketing dataset contains `r nrow(marketing)` observations across `r ncol(marketing)` variables. The variables include numeric fields like marketing spend and purchase amount, as well as categorical fields like customer segment and notes.

```{r}
#| label: tbl-segment-summary
#| tbl-cap: "Sample size and mean values of key numeric variables by customer segment."
#| echo: false

marketing |>
  group_by(segment) |>
  summarise(
    `N`                    = n(),
    `Mean spend (\\$)`     = round(mean(marketing_spend), 2),
    `Mean visits`          = round(mean(visits), 2),
    `Mean purchase (\\$)`  = round(mean(purchase_amount), 2),
    `Median purchase (\\$)` = round(median(purchase_amount), 2)
  ) |>
  kable()
```

As shown in @tbl-segment-summary, the majority of customers belong to the Consumer segment, while the Enterprise segment has very few customers.
@fig-segment-notes shows that most customers in all segments did not have any outreach notes recorded.

```{r}
#| label: fig-segment-notes
#| fig-cap: "Distribution of outreach status (notes) within each customer segment. The majority of customers have no note recorded across all segments."
#| fig-width: 7
#| fig-height: 2.8
#| echo: false

marketing |>
  ggplot(aes(x = segment, fill = notes)) +
  geom_bar(position = "stack") +
  scale_fill_brewer(palette = "Set2") +
  labs(x = "Customer segment",
       y = "Number of customers",
       fill = "Notes") +
  theme_minimal(base_size = 11)
```

# Results

```{r}
#| label: fig-spend-purchase
#| fig-cap: "Purchase amount against marketing spend for all customers. The blue line is a linear fit; the shaded band is its 95% confidence interval."
#| fig-width: 6.5
#| fig-height: 2.8
#| echo: false

marketing |>
  ggplot(aes(x = marketing_spend, y = purchase_amount)) +
  geom_point(alpha = 0.55, colour = "steelblue") +
  geom_smooth(method = "lm", se = TRUE, colour = "darkblue") +
  labs(x = "Marketing spend per customer ($)",
       y = "Purchase amount ($)") +
  theme_minimal(base_size = 11)
```

```{r}
#| label: fig-spend-purchase-segment
#| fig-cap: "Purchase amount against marketing spend, split by customer segment. Each panel shows its own linear fit."
#| fig-width: 7.5
#| fig-height: 3.0
#| echo: false

marketing |>
  ggplot(aes(x = marketing_spend, y = purchase_amount,
             colour = segment)) +
  geom_point(alpha = 0.65) +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~ segment) +
  scale_colour_brewer(palette = "Dark2") +
  labs(x = "Marketing spend per customer ($)",
       y = "Purchase amount ($)",
       colour = "Segment") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "none")
```

```{r}
#| label: results-stats
#| echo: false

r_overall <- cor(marketing$marketing_spend, marketing$purchase_amount)
fit       <- lm(purchase_amount ~ marketing_spend, data = marketing)
slope     <- coef(fit)[2]
r_squared <- summary(fit)$r.squared

by_segment <- marketing |>
  group_by(segment) |>
  summarise(r = cor(marketing_spend, purchase_amount),
            n = n())

r_consumer   <- by_segment$r[by_segment$segment == "Consumer"]
r_smb        <- by_segment$r[by_segment$segment == "SMB"]
r_enterprise <- by_segment$r[by_segment$segment == "Enterprise"]
n_enterprise <- by_segment$n[by_segment$segment == "Enterprise"]
```

Based on the data, there is a strong positive relationship between marketing spend and how much a customer purchases, as shown in @fig-spend-purchase. Overall, the correlation is `r round(r_overall, 2)`, meaning that as marketing spend increases, the purchase amount tends to increase as well. Specifically, for every extra dollar spent on marketing, the purchase amount increases by about `r round(slope, 2)` dollars. When we look at the data split by group in @fig-spend-purchase-segment, this positive trend is very clear for both the Consumer and SMB segments. The Enterprise segment also shows a positive trend, but the data are much more spread out, likely because there are very few customers in that group. Overall, to answer the research question, higher marketing spend is clearly associated with higher purchase amounts, suggesting that the personalised marketing campaign was effective.